A Reimplementation of Python’s dict using Simple Tabulation Hashing and Linear Probing

### 2012-05-20 docs/paper.tex
Explain hash table shootout

For simple tabulation hashing with variable length strings, we decided to use a
fixed number of tables $N$ and round robin between them for every byte.  We
found the optimal value of $N$ to be 8 experimentally.  We limited our choices
to powers of 2 so that costly modulo operations could be avoided by using bit
masks Figure ~\ref{fig:tables} shows the performance of different numbers of
tabulation tables with both 1 and 2 byte indices.  Eight hash tables indexing
every byte separately resulted in the best performance.    We believe this is due
to the low entropy for fewer tables and too much memory bandwidth for a higher
number of tables.  The optimal result may also be an artifact of the fact that
we were using on the order of magnitude of thousands to millions of dictionary
keys.  Our optimal result resulted 256 64
bit table entries in each of 8 tables using 16KB of memory.
 \begin{figure}[H]
   \centering
   \includegraphics[width=4.1in]{tables.pdf}
   \caption{Eight tabulation tables with 1 byte indices provides the best performance for a
   dictionary with 6000000 keys}
   \label{fig:tables}
 \end{figure}

 \begin{figure}[H]
   \centering
   \includegraphics[width=4.1in]{randstring.pdf}
   \caption{Random string performance}
   \label{fig:randstring}
 \end{figure}
 \begin{figure}[H]
   \centering
   \includegraphics[width=4.1in]{seqstring.pdf}
   \caption{Sequential string performance}
   \label{fig:seqstring}
 \end{figure}
 \begin{figure}[H]
   \centering
   \includegraphics[width=4.1in]{chainlength.pdf}
   \caption{Simple tabulation hashing decreases the average chain length}
   \label{fig:seqstring}
 \end{figure}

\subsection{Benchmarks}
In order to analyze the result of this new dictionary we measured three main metrics. First, we compared the performance of the simple tabulation hashing with Python’s current hashing scheme on random strings. Second, we measured the number of probes and collisions and overall the quality of our dictionary on structured and unstructured inputs. Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
Because of the heavy utilization of dictionaries in Python internals one of the valuable benchmarks is to see how the new dictionary performs on them. We used PyBench benchmark suit for this task. We saw significant improvements of around \%20 in performance of some of the internals such as the special and normal class attribute lookups and string predicates. However, at the same time some of the other internals such as CPython exception handler were running around \%12 slower compared to the current implementation. On average a \%4.5 improvement in runtime was measures for running the whole benchmark suite.

@@

In a couple of the programs such as pure Python Pickle (an object serialization library) and html5lib (specification parser library) we calculated an improvement of \%3 to \%4. In other cases we did not calculate an speedup but at the same time we did not find any real-world case which the new implementation performs significantly slower. These results were consistent with the previous results of PyBench that showed about \%4.5 improvement.

>>>>

For simple tabulation hashing with variable length strings, we decided to use a
fixed number of tables $N$ and round robin between them for every byte.  We
found the optimal value of $N$ to be 8 experimentally.  We limited our choices
to powers of 2 so that costly modulo operations could be avoided by using bit
masks Figure ~\ref{fig:tables} shows the performance of different numbers of
tabulation tables with both 1 and 2 byte indices.  Eight hash tables indexing
every byte separately resulted in the best performance.    We believe this is due
to the low entropy for fewer tables and too much memory bandwidth for a higher
number of tables.  The optimal result may also be an artifact of the fact that
we were using on the order of magnitude of thousands to millions of dictionary
keys.  Our optimal result resulted 256 64
bit table entries in each of 8 tables using 16KB of memory.
 \begin{figure}[H]
   \centering
   \includegraphics[width=4.1in]{tables.pdf}
   \caption{Eight tabulation tables with 1 byte indices provides the best performance for a
   dictionary with 6 million keys}
   \label{fig:tables}
 \end{figure}

\subsection{Synthetic Benchmarks}

Our primary benchmark was a synthetic one that called the CPython dictionary
code directly from C so that all the interpreter overhead was bypassed.  We
measured performance for both sequential strings, generated by converting
sequential integers into their string representation, and for random strings,
generated by the string representation of random integers.  For each of the
benchmarks, we inserted a given number of keys into the dictionary and measured
the runtime.

Figure ~\ref{fig:randstring} shows the performance for random strings.  Both
linear probing and simple tabulation result in improved performance over the
current CPython dictionary implementation but combining both techniques results
in over a 20\% improvement.  Figure ~\ref{fig:seqstring} shows very similar
results for sequential strings.

\begin{figure}[H]
  \centering
  \includegraphics[width=4.1in]{randstring.pdf}
  \caption{Random string performance}
  \label{fig:randstring}
\end{figure}
\begin{figure}[H]
  \centering
  \includegraphics[width=4.1in]{seqstring.pdf}
  \caption{Sequential string performance}
  \label{fig:seqstring}
\end{figure}

Figure ~\ref{fig:chainlength} shows the average chain length, or dictionary
cell probes per lookup for a run of the benchmark on sequential strings.
Simple tabulation hashing with CPython's current collision resolution strategy
results in the fewest number of probes.  However, despite having the fewest
number of probes, its real relative performance is much worse as seen in Figure
~\ref{fig:seqstring}.  Linear probing performs faster than its average
chain length suggests, showing the impact of looking at entries in the same
cache line first rather than jumping around.

\begin{figure}[H]
  \centering
  \includegraphics[width=4.1in]{chainlength.pdf}
  \caption{Simple tabulation hashing decreases the average chain length}
  \label{fig:chainlength}
\end{figure}

\subsection{Benchmarks}
In order to analyze the result of this new dictionary we measured three main metrics. First, we compared the performance of the simple tabulation hashing with Python’s current hashing scheme on random strings. Second, we measured the number of probes and collisions and overall the quality of our dictionary on structured and unstructured inputs. Finally, in this section  we show some of the total time spent in dictionary operations by using a set of general Python benchmarks and some real-world programs.
Because of the heavy utilization of dictionaries in Python internals one of the
valuable benchmarks is to see how the new dictionary performs on them. We used
PyBench benchmark suit for this task. We saw significant improvements of around
20\% in performance of some of the internals such as the special and normal
class attribute lookups and string predicates. However, at the same time some
of the other internals such as CPython exception handler were running around
12\% slower compared to the current implementation. On average a 4.5\% improvement in runtime was measures for running the whole benchmark suite.

@@

In a couple of the programs such as pure Python Pickle (an object serialization
library) and html5lib (specification parser library) we calculated an
improvement of 3\% to 4\%. In other cases we did not calculate an speedup but
at the same time we did not find any real-world case which the new
implementation performs significantly slower. These results were consistent
with the previous results of PyBench that showed about 4.5\% improvement.
